Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-12-21 Paper.Rmd
further edits

Question Q1 was answered principally via analysis of the
data using a model selection approach. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. Ranks were $1$ for the smallest element in a set. 
Second, we did a test of the independence of the $u_i$ and $v_i$ (implemented in X in the Y package). 
<!--***DAN: X and Y to be flled in, can't remember right now what they are and don't want to stop
to look it up-->
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. For keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

We also defined the statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b_k},u_{b_k}}(u,v)$; we use both statistics 
because in some instances $\Ps_{l_b,u_b}$ appears to have more power to
reveal tail dependence than does $\cor_{l_{b_k},u_{b_k}}(u,v)$. 
<!--***DAN: need to make this a bit more precise, perhaps just enough to cite one of the supp mat sections where tests are done of these stats, but you need to review the tests first and what they show-->
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
<!--***DAN: Shyamolina, is this actually right, i.e., did we use the average, or did we 
use the sum divided by n? Whichever way we did it is OK, but please check the code and revise the above 
text if necessary and insert a comment answering this comment. I tried to check this, myself, 
but I was having a hard time finding where the statistics are implemented. Please also
indicate in your response comment what line of what file has the implementation of these
statistics.
***Shya: In code we used the sum of squared distances divided by n(=number of points lying within two boundaries) which I think okay to be called as "average-squared distance" or (to be precised) average of squared distances of points from 1:1 line.
Reference : CopulaFunctions_flexible.R file, line 24.-->
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, which are large when points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ cluster close to the line $v=u$, $\Dsq_{l_b,u_b}$ is 
small in that case. 
Thus large values of $\cor_{l_b,u_b}$ and $\Ps_{l_b,u_b}$ indicate strong 
dependence in the portions of the distributions of $u$ and $v$ given by 
the bounds, whereas small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

Across essentially all the sub-fields and levels of organization within ecology that we 
examined, copula structures were typically non-normal and variables exhibited asymmetric tail 
dependence. To make results easier for readers who are new to copulas to absorb: we first 
present results in detail for an example bivariate dataset; followed by a summary of results
for all bivariate datasets; then we present results in 
detail for an example multivariate dataset; followed by a summary of results 
for all multivariate datasets.

<!--bivariate example-->
We will now discuss briefly findings from one of our bivariate datasets as follows. As described in the \nameref{Data}, raw data plot for RaCA database soil organic C and soil N stocks to 1m of soil depth from the central pedon of the `r nrow(d)` sites and their corresponding copula plot were presented in Fig. \ref{fig_biv_multi_raw_cop}(A) and (E), respectively.
We carried out model selection analysis and non-parametric statistics to measure tail-dependence (if any) on this copula plot.

<!--#### Model selection approach \label{Ex_SoilCN_Model_selection}-->

From the model selection approach, the test of indepedence gave $p=$ `r BivMS_res_Loecke$IndepTestRes` (to within the precision available from `BiCopIndTest` in the `VineCopula` package) that showed a significant correlation (Kendall's coefficient = `r round(BivMS_res_Loecke$TauVal,2)`). We then fitted several copulae (all $16$ families as mentioned in \nameref{Model_selection})
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulae (Table \ref{SM-tab_soilCNfit}). 
Based on AIC weights, `r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the best fit. The table shows that several copulae are much better supported (higher AIC and BIC weights) than the Normal copula, answering the first part of \hyperref[Intro]{\textbf{(Q1)}} for these data.

The model-averaged (AIC weights used for model averaging) lower-tail dependence statistic was `r BivMS_res_Loecke$relLTdep_AICw` and the model-averaged upper-tail dependence statistic was `r BivMS_res_Loecke$relUTdep_AICw`. These values are distinct from what a Normal copula would give (i.e., 0), answering the second part of \hyperref[Intro]{\textbf{(Q1)}} for these data. Although these results do convincingly show non-Normal copula structure, we should caution that because even the lowest-AIC copula was a poor fit, giving $p$-value `r BivMS_res_Loecke$GofRes_CvM` (to within the precision available from `BiCopGofTest` in the `VineCopula` package) in the Cramer-von Mises-based test of the goodness of its fit, and $p$-value `r BivMS_res_Loecke$GofRes_KS` in the Kolmogorov-Smirnov-based test of the goodness of fit. Therefore, we also verified this finding with \nameref{NPA} approach (see Fig. \ref{SM-fig_soilCN_nonparam}), where each of our statistics ($\cor$, $\Ps$, $\Dsq$), as computed for the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (with $10$ number of bins), was compared to the distributions of values of the same statistic with the same ranges computed on 1000 Normal-copula surrogate datasets with similar Kendall or Spearman correlations, in different runs for Kendall and Spearman (Fig. \ref{SM-fig_soilCN_nonparam}). Each bin in each subplot of Fig. \ref{SM-fig_soilCN_nonparam} contains a number indicating how many Normal surrogates out of 1000, are different ($95\%$ confidence interval) from Normal-copula statistics. Those numbers clearly show that tail dependence of data is stronger in both the lower and upper tails than would be expected under a null hypothesis of a normal copula, answering the second part of \hyperref[Intro]{\textbf{(Q1)}}.

>>>>

Question Q1 was answered principally via analysis of the
data using a model selection approach. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. Ranks were $1$ for the smallest element in a set. 
Second, we did a test of the independence of the $u_i$ and $v_i$ (implemented in X in the Y package). 
<!--***DAN: X and Y to be flled in, can't remember right now what they are and don't want to stop
to look it up, Shya, you could also fill these in if you get a chance-->
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. For keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

We also defined the statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b_k},u_{b_k}}(u,v)$; we use both statistics 
because in some instances $\Ps_{l_b,u_b}$ appears to have more power to
reveal tail dependence than does $\cor_{l_{b_k},u_{b_k}}(u,v)$. 
<!--***DAN: need to make this a bit more precise, perhaps just enough to cite one of the supp mat sections where tests are done of these stats, but you need to review the tests first and what they show-->
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
<!--***DAN: Shyamolina, is this actually right, i.e., did we use the average, or did we 
use the sum divided by n? Whichever way we did it is OK, but please check the code and revise the above 
text if necessary and insert a comment answering this comment. I tried to check this, myself, 
but I was having a hard time finding where the statistics are implemented. Please also
indicate in your response comment what line of what file has the implementation of these
statistics.
***Shya: In code we used the sum of squared distances divided by n(=number of points lying within two boundaries) which I think okay to be called as "average-squared distance" or (to be precised) average of squared distances of points from 1:1 line.
Reference : CopulaFunctions_flexible.R file, line 24.
***DAN: OK, good, I am leaving these comments here as a reminder-->
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, which are large when points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ cluster close to the line $v=u$, $\Dsq_{l_b,u_b}$ is 
small in that case. 
Thus large values of $\cor_{l_b,u_b}$ and $\Ps_{l_b,u_b}$ indicate strong 
dependence in the portions of the distributions of $u$ and $v$ given by 
the bounds, whereas small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

Across essentially all the sub-fields and levels of organization within ecology for which we 
examined a dataset, copula structures were typically non-normal and variables 
commonly exhibited asymmetric tail 
dependence. To make results easier to absorb for readers who are new to copulas: we first 
present results in detail for an example bivariate dataset; followed by a summary of results
for all bivariate datasets; then we present results in 
detail for an example multivariate dataset; followed by a summary of results 
for all multivariate datasets.

<!--bivariate example-->
We discuss the example of the soil C and N data (see Data), pictured in 
Fig. \ref{fig_biv_multi_raw_cop}A (raw data) and E (copula plot). The variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent
(Fig. \ref{fig_biv_multi_raw_cop}E).
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted the 16 copulas
mentioned in the previous section to the 
data pictured in Fig. \ref{fig_biv_multi_raw_cop}E; AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions were computed (Table \ref{SM-tab_soilCNfit}). 
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
normal copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
The model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for model averaging). 
These values are distinct from what a normal copula would give, namely, 0, answering the second part 
of Q1 for these data. The values are also substantially 
distinct from each other, answering the third part of Q1.

<!--***DAN: made it to here 2018 12 20 920pm-->

Although these results do convincingly show non-normal copula structure, 
we caution that the model-averaged tail-dependence statistics may be biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest` in the `VineCopula` package). However, nonparametric statistics verified
both the asymmetry of tail dependence for the soil data and the conclusion that tail dependence 
in the data was distinct from a normal copula (Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}).

<!--***DAN: the above block of text is the first paart of a para, and i have edited it. The below
block is the rest of the same para, and I have not-->

where each of our statistics ($\cor$, $\Ps$, $\Dsq$), as computed for the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (with $10$ number of bins), was compared to the distributions of values of the same statistic with the same ranges computed on 1000 Normal-copula surrogate datasets with similar Kendall or Spearman correlations, in different runs for Kendall and Spearman (Fig. \ref{SM-fig_soilCN_nonparam}). Each bin in each subplot of Fig. \ref{SM-fig_soilCN_nonparam} contains a number indicating how many Normal surrogates out of 1000, are different ($95\%$ confidence interval) from Normal-copula statistics. Those numbers clearly show that tail dependence of data is stronger in both the lower and upper tails than would be expected under a null hypothesis of a normal copula, answering the second part of \hyperref[Intro]{\textbf{(Q1)}}.

### 2018-12-21 SupportingInformation.Rmd
further edits

<!--Display model selection results in a table-->
```{r tab_soilCNfit, echo=F, results="markup"}
#knitr::opts_knit$set(kable.force.latex = TRUE)
h<-BivMS_res_Loecke$InfCritRes[,c("copname","AICw","BICw","LTdep","UTdep")]
knitr::kable(h, #longtable = T, 
             format = "latex",
      caption.short = "Fitting info for soil C and N dataset",
      caption = "Fitting information for **soil C and N dataset**. Here and henceforth copula names are abbreviated as: N=normal (Gaussian); C=Clayton; G=Gumbel; F=Frank; J=Joe; SC=survival Clayton; SG=survival Gumbel; SJ=survival Joe. An S in front of one of the copulas BB1, BB6, BB7, BB8 indicates the survival version of that copula (rotation by $180^\\circ$).\\label{tab_soilCNfit}",
      booktabs = TRUE)
```

>>>>

<!--Display model selection results in a table-->
```{r tab_soilCNfit, echo=F, results="markup"}
#knitr::opts_knit$set(kable.force.latex = TRUE)
h<-BivMS_res_Loecke$InfCritRes[,c("copname","AICw","BICw","LT","UT")]
knitr::kable(h, #longtable = T, 
             format = "latex",
      caption.short = "Model fitting results for soil C and N dataset",
      caption = "Results of fitting our 16 copulas to the soil C and N dataset. Copula name abbreviations are given in the main text. AICw = AIC weight; BICw = BIC weight; LT = the lower-tail dependence statistic for the copula family with the fitted parameters; UT = the same for upper-tail dependence.\\label{tab_soilCNfit}",
      booktabs = TRUE)
```
